Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-05 paper.qmd
almost there...

The recent decades observed an increase in the number of space activities.
This resulted in higher numbers of re-entries at the end of the respective mission
lifespan or orbital decay. The risk of collision between uncontrolled
re-entering space debris and aircraft is increasing due to the continual 
growth of flights and of orbiting bodies and their relevant re-entries). 
<!-- why important --> National and international
authorities need to balance safety and economic impact of adopted
measures. This is particularly challenging should a re-entry event be announced on short notice.
Thus, there is a need for effective tools to reduce the uncertainty and increase the safety awareness of uncontrolled re-entries. 
<!-- approach / experiments --> To determine the risk exposure, this paper follows a data-driven approach.
We employ hourly and daily spatial flight densities calculated from trajectories as recorded by
EUROCONTROL's Network Manager and binned with the Uber H3 geospatial
indexing system. Impact risk probabilities are then calculated based on previous
theoretical work. 
<!-- results / conclusions --> This work enhances the
model by using actual flown flight trajectories as a basis for
time-varying flight densities rather than relying on proxies such as the
population density or ADS-B/Mode S message counts. These densities
combined with a buffered area along the eventual path of re-entry, can
provide mid-air impact probabilities to authorities as an aid for
deciding on safety measures such as airspace closures and/or
re-routings.
The findings and the approach are demonstrated on the basis of an expected re-entry from an earlier space mission.\
       IS IT WORTH TO STATE THE DETERMINED RISK?  \
The approach in this paper enables a computing time efficient mechanism to support the re-entry related decision-making.

@@

<!-- next sentence feels like an orphan ... do we need this here? ------>
The next sections will develop the mathematical background for the
evaluation of the collision and casualty risk for which air traffic density is
a key component.

@@

Our approach builds on the discretisation of the geospatial risk area.
For this purpose we use a hexagon-based tessellation pattern.
Uber's H3 method ensures non-overlapping hexagons (and 5 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
While this paper focusses on the determination of the risk exposure per cell, the approach would allow to identify adjacent cells to account for the inherent uncertainty and risk appetite of decision-makers.

@@

Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes (i.e., flight levels).
The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
At the same time, the trajectory of a falling re-entering debris object at regular en-route flight levels will be a vertical decent movement[^ballistic_descent].
This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as

@@

* constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$ 
  ([we adopt this value from pag 7-8 in @patera2008])
* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, under the
  simplification that the majority of the flight time spent cruising, 
* aircraft area as viewed from the top, $a_T = W_t \times L_t = T_t$, 
  where $W_t$ is the wingspan, and $L_t$ the length for the specific aircraft
  type $t$
* aircraft area as viewed from the front, $a_F = W_t \times H_t = F_t$, 
  where $W_t$ is the wingspan, and $H_t$ the height for the specific aircraft
  type $t$

@@

So for example if a building is located at latitude 30 degrees and has an area 
of $\qty{1000}{m^2}$ and a piece of debris re-enters the Earth’s atmosphere
from an orbit having an inclination of 45 degrees, the impact probability is 
given by

@@

$$ \sigma(\varphi) = \sum_k \sigma(\varphi, i_k)$$

@@

Spatially projecting $\sigma(\lambda,\varphi)$ on the H3 cells in the
\acr{NM} area, we obtain the weightings per cell (per hour $H$), $W_H(h)$, as in
@fig-weighting-nm-area.

@@

Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of re-entry.
The densest cell, `83194afffffffff` (area \qty{11131585027}{m^2}), at 24.56 flights/hour has centroid latitude
of \ang{50.94369} so within the latitude range of $[\ang{-51.7}, \ang{51.7}]$.
Using an average value for effective exposed area of \qty{1000}{m^2},
we can quickly calculate the collision expectation as

$$
\begin{aligned}
E &= w_h \cdot \sigma_h \cdot \bar{\varepsilon} / area_h \\
  &= 10^{-6} \cdot 24.56 \cdot 1000 / 11131585027 = 2.2\cdot 10^{-12}
\end{aligned}
$$

@@

We have calculated collision and casualty risk probabilities for the two
busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
as logically depicted in @fig-data-densities is available in the companion
GitHub repository of the paper at <https://github.com/euctrl-pru/flight_density_space_debris>.
We have also developed a companion R package `aviodebris` to heavy-lift the
various data processing steps to produce the results in the paper;
it is available at <https://github.com/euctrl-pru/aviodebris>.

::: {#fig-data-densities}
```{r densities}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/cubes.png")
```

@@

This paper expands on previous work. With a focus on Europe, we
establish an upper bound for the traffic density based on actual
flights.

... SOME MORE STUFF DONE IN THE PAPER ...

```         
    WHAT ARE THINGS THAT ARE "WEAKNESSES" = STUFF THAT CAN BE IMPROVED? IT WOULD BE GOOD TO HAVE A LINE OR TWO ON THIS.
```

Future work will address the shortcomings.
<!-- or can we say something about the next concrete steps to remedy some weaknesses -->

::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h"}
```{r ellipse-parameters}
#| fig-cap: "Keplerian elements: semi-major axis, $a$, the distance between the foci, $2c$, the eccentricity, $e = (2c)/(2a)$. (the semi-parameter (a.k.a. _semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as $p = b^2 / a = a (1 - e^2)$)"

@@

>>>>

The recent decades observed an increase in the number of space activities.
This resulted in higher numbers of re-entries at the end of the respective mission
lifespan or orbital decay. The risk of collision between uncontrolled
re-entering space debris and aircraft is increasing due to the continual 
growth of flights and of orbiting bodies and their relevant re-entries. 
<!-- why important --> National and international
authorities need to balance safety and economic impact of adopted
measures. This is particularly challenging should a re-entry event be announced on short notice.
Thus, there is a need for effective tools to reduce the uncertainty and increase the safety awareness of uncontrolled re-entries. 
<!-- approach / experiments --> To determine the risk exposure, this paper follows a data-driven approach.
We employ hourly and daily spatial flight densities calculated from trajectories as recorded by
EUROCONTROL's Network Manager and binned with the Uber H3 geospatial
indexing system. Impact risk probabilities are then calculated based on previous
theoretical work. 
<!-- results / conclusions --> This work enhances the
model by using actual flown flight trajectories as a basis for
time-varying flight densities rather than relying on proxies such as the
population density or ADS-B/Mode S message counts. These densities
combined with a buffered area along the eventual path of re-entry, can
provide mid-air impact probabilities to authorities as an aid for
deciding on safety measures such as airspace closures and/or
re-routings.
The findings and the approach are demonstrated on the basis of an expected re-entry from an earlier space mission.\
       IS IT WORTH TO STATE THE DETERMINED RISK?  \
The approach in this paper enables a computing time efficient mechanism to support the re-entry related decision-making.

@@

<!-- next sentence feels like an orphan ... do we need this here? ------>
The next sections will develop the mathematical background for the
evaluation of the collision risk for which air traffic density is
a key component.

@@

Our approach builds on the discretisation of the geospatial risk area.
For this purpose we use a hexagon-based tessellation pattern.
Uber's H3 method ensures non-overlapping hexagons (and 12 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
While this paper focusses on the determination of the risk exposure per cell, the approach would allow to identify adjacent cells to account for the inherent uncertainty and risk appetite of decision-makers.

@@

Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes (i.e., flight levels).
The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
At the same time, the trajectory of a falling re-entering debris object at regular en-route flight levels will be a vertical descent movement[^ballistic_descent].
This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as

@@

* constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$ 
  ([we adopt this value from pag 7-8 in @patera2008])
* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, under the
  simplification that the majority of the flight time is spent cruising, 
* aircraft area as viewed from the top, $a_T = W_t \times L_t = T_t$, 
  where $W_t$ is the wingspan, and $L_t$ the length for the specific aircraft
  type $t$
* aircraft area as viewed from the front, $a_F = W_t \times H_t = F_t$, 
  where $W_t$ is the wingspan, and $H_t$ the height for the specific aircraft
  type $t$

@@

So for example if a building is located at latitude \ang{30} and has an area 
of $\qty{1000}{m^2}$ and a piece of debris re-enters the Earth’s atmosphere
from an orbit having an inclination of \ang{45}, the impact probability is 
given by

@@

$$ \sigma(\varphi) = \frac{1}{N} \sum_{k = 1}^N \sigma(\varphi, i_k)$$

@@

Spatially projecting $\sigma(\lambda,\varphi)$ on the H3 cells in the
\acr{NM} area, i.e. taking the mean of all the half degree weighting values
and spreading in longitude,
we obtain the weightings per cell (per hour $H$), $W_H(h)$, as in
@fig-weighting-nm-area.

@@

::: {#fig-kosmos-cell}
```{r kosmos-cell}
#| out-width: "45%"
#| echo: false
knitr::include_graphics(here::here("media", "figures", "cell_london.png"))
```

Cell with highest collision expectation due to KOSMO debris re-entry based on 2024-07-05 traffic.
:::

Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of re-entry.
Using the approach in this study and the weightings for this specific debris,
the cell with the highest impact risk is cell, `83194afffffffff` (located South
of London, cf. @fig-kosmos-cell), with a value of $1.89 \cdot  10^{-10}$.

@@

We have calculated collision risk probabilities for the two
busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
as logically depicted in @fig-data-densities.

::: {#fig-data-densities}
```{r densities}
#| out-width: "80%"
#| echo: false
knitr::include_graphics("media/figures/cubes.png")
```

@@

Code and sample data is available in the companion GitHub repository of
the paper at <https://github.com/euctrl-pru/flight_density_space_debris>.
We have also developed an R package `aviodebris` to heavy-lift the
various data processing steps to produce the results in the paper;
it is available at <https://github.com/euctrl-pru/aviodebris>.

::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="htb"}
```{r ellipse-parameters}
#| fig-cap: "Keplerian elements: semi-major axis, $a$, the distance between the foci, $2c$, the eccentricity, $e = (2c)/(2a)$. (the semi-parameter (a.k.a. _semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as $p = b^2 / a = a (1 - e^2)$)"

@@

This paper expands on previous work. With a focus on Europe, we
establish an upper bound for the traffic density based on actual
flights.
In particular we derive impact risk probability for the European airspace
using flown trajectories.

Future work will address the derivation of casuality expectation using
seats information for aircraft types in the traffic.
<!-- or can we say something about the next concrete steps to remedy some weaknesses -->
